---
title: "Milestone 3- Data Preparation & Analysis"
author: "Kaylynn Mosier"
date: "2024-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Population Change Dataset

## Data Preparation

```{r}
# Load libraries
# plot based on percentage change greater than 25%
library(ggplot2)
library(dplyr)
library(readxl)
library(tidyr)
library(stringr)
```

```{r}
# open excel file
population_data <- read_excel("C:/Users/kayly/OneDrive/Desktop/MSDS/DSC630/Final Project/Population Change.xlsx",col_names = FALSE)

# Drop City column
population_data <- population_data[,-5]

# Set column titles
names(population_data) <- c("City_State", "Population_2020","Population_2021", 
                            "Population_2022", "Population_2023")

# Delete rows 1 through 4, they were multi-level column titles in original dataset
population_data <- population_data[-c(1:4),]
```

```{r}
# Check number of rows
nrow(population_data)
```

```{r}
# Remove last 4 rows, they include citation information
population_data <- population_data[-c(19486:19490),]
```

```{r}
# Split City_State column into City and State columns
population_data[c('City', 'State')] <- str_split_fixed(population_data$City_State, ',', 2)

# Drop City_State column, it is no longer needed
population_data <- population_data[, -1]
```

```{r}
# Remove 'city' and 'town' from city names
library(tm)
stopwords <- c("city", "town")  # Words to remove
x = population_data$City # Storing column to search in as variable
x = removeWords(x, stopwords)  # Remove stop words from column

# Add City column with stop words removed as new column
population_data$City_cleaned <- x  

# Drop City column
population_data <- population_data[,-5]
```

```{r}
# Check number of NAs
colSums(is.na(population_data))
```
```{r}
# Drop rows that contain NAs
population_data <- population_data[complete.cases(population_data), ]

# Confirm rows with NA were dropped
colSums(is.na(population_data))
```


```{r}
# Change Population_2020 to numeric datatype
population_data[, 1] <- sapply(population_data[,2], as.numeric)

# Confirm data type changed
sapply(population_data, class)
```

```{r}
# Add population change column
population_data <- population_data %>%
    mutate(popChange_2020_to_2023 = Population_2020 - Population_2023)
```

## Visualizations
```{r}
# Plot average population change by state
pop_change_data <- population_data %>%
    group_by(State) %>%
    summarize(Mean_Change = mean(popChange_2020_to_2023)) %>%
    ggplot(aes(x = reorder(State, Mean_Change), y = Mean_Change)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    coord_flip() +
    labs(title = "Average Population Change by State",
         x = "State",
         y = "Average Population Change") +
    theme(text = element_text(size = 6))
print(pop_change_data)
```

```{r}
# Find top 10 cities with the highest increase in population
# Plot top 10 cities with highest increase in population
top_10 <- population_data %>%
    arrange(desc(popChange_2020_to_2023)) %>%
    slice(1:10) %>%
    ggplot() +
    geom_bar(aes(x=City_cleaned, y=popChange_2020_to_2023), 
             stat='identity', fill='steelblue') +
    labs(title = "Top 10 Population Changes by City",
         x = "City",
         y = "Population Change") +
    theme(axis.text.x = element_text(angle = 45, hjust=1)) +
    scale_y_continuous(labels = scales::comma)

top_10

```

# Restaurant Locations Data
## Data Preparation

```{r}
# Load data
restaurant_locations <- read_excel("C:/Users/kayly/OneDrive/Desktop/MSDS/DSC630/Final Project/restaurant_locations_final.xlsx")
```

```{r}
# Remove dateAdded, dateUpdated, address, country, keys, sourceURLs, website, and postalCode columns
restaurant_locations <- restaurant_locations %>%
    select(-c(dateAdded, dateUpdated, address, country, keys, 
              sourceURLs, websites, postalCode))
```


```{r}
# Check for missing values
colSums(is.na(restaurant_locations))
```
```{r}
# Remove row that contains NA
restaurant_locations <- restaurant_locations[complete.cases(restaurant_locations), ]

# Confirm all NAs were dropped
colSums(is.na(restaurant_locations))
```


```{r}
length(unique(restaurant_locations$categories))
```
There are way too many categories to do anything with. Looking at the data shows many variations of the same few categories (ex: Fast Food, Fast Food Restaurant, etc). These need to be fixed to reduce the number of categories

```{r}
library(stringr)

# Check if categories contain 'Fast Food'
restaurant_locations$Fast_Food <- 
    grepl('FAST FOOD', toupper(restaurant_locations$categories), fixed=TRUE)

# Check unique values in Fast_Food column
unique(restaurant_locations$Fast_Food)
```
All columns contain 'Fast Food', so all restaurants are fast food restaurants. This column does not add value to analysis so it can be dropped

```{r}
restaurant_locations <-
    restaurant_locations %>% 
    select(-c(categories, Fast_Food))
```


```{r}
# Create dataframe of state name and state abbreviation
province <- state.abb
state <- state.name
state_abb_name <- data.frame(province, state)

# Join state_abb_name to restaurant_locations on province column
restaurant_locations <- merge(restaurant_locations, state_abb_name, by='province')

# Drop province column from restaurant_locations
restaurant_locations <- restaurant_locations %>%
    select(-c(province))
```

## Visualizations


# QSR Top 50 Data
```{r}
# Load data
QSR_top_50_2022 <- read_excel("C:/Users/kayly/OneDrive/Desktop/MSDS/DSC630/Final Project/QSR_top_50.xlsx",sheet= 1)
QSR_top_50_2021 <- read_excel("C:/Users/kayly/OneDrive/Desktop/MSDS/DSC630/Final Project/QSR_top_50.xlsx",sheet= 2)
```

```{r}
# Add column for year to each QSR dataset
QSR_top_50_2021$Year <- c(2021)
QSR_top_50_2022$Year <- c(2022)
```

```{r}
# Change column name, so titles are the same in both datasets
QSR_top_50_2021 <- rename(QSR_top_50_2021, 
                          'Rank'='2021 Rank',
                          'System_Sales_Millions'='2021 US Systemwide Sales Millions',
                          'Avg_Sales_Per_Unit_Thousands'='2021 Average Sales Per Unit Thousands', 
                          'Franchised_Units'='2021 Franchised Licensed Units',
                          'Company_Units'='2021 Company Units',
                          'Total_Units'='2021 Total Units',
                          'Total_Change_From_Previous_Year' ='Total Change in Units from 2020')

QSR_top_50_2022 <- rename(QSR_top_50_2022, 
                          'Rank'= '2022 Rank',
                          'System_Sales_Millions'='2022 US Systemwide Sales Millions',
                          'Avg_Sales_Per_Unit_Thousands'='2022 Average Sales Per Unit Thousands', 
                          'Franchised_Units'='2022 Franchised Licensed Units',
                          'Company_Units'='2022 Company Units',
                          'Total_Units'='2022 Total Units',
                          'Total_Change_From_Previous_Year' ='Total Change in Units from 2021')
```


```{r}
# Combine both QSR datasets into one
QSR_top_50 <- merge(QSR_top_50_2021, QSR_top_50_2022, 
                    by=c('Company', 'Year', 'Category','Rank','System_Sales_Millions',
                         'Avg_Sales_Per_Unit_Thousands', 'Franchised_Units', 'Company_Units',
                         'Total_Units','Total_Change_From_Previous_Year'),all=TRUE)
```

```{r}
# Check number of NAs
colSums(is.na(QSR_top_50))
```
The two datasets are now fully merged and can be filtered by year.

## Visualizations

```{r}
subset_QSR_2022 <- subset(QSR_top_50, Year==2022)
# Plot Total_change_From_Previous_Year for each restaurant in each year
unit_change_from_previous_year <- subset_QSR_2022 %>%
    ggplot(aes(x=reorder(Company, Total_Change_From_Previous_Year),
               y= Total_Change_From_Previous_Year))+
    geom_bar(stat="identity", fill='steelblue') +
    coord_flip() +
    labs(title = "Unit Change from 2021 to 2022 by Company",
         x = "Company",
         y = "Unit Change From Previous Year") +
    theme(text = element_text(size = 6))
    

unit_change_from_previous_year
```
```{r}
subset_QSR_2021 <- subset(QSR_top_50, Year==2021)

# Plot Total_change_From_Previous_Year for each restaurant in each year
unit_change_2020_to_2021 <- subset_QSR_2021 %>%
    ggplot(aes(x=reorder(Company, Total_Change_From_Previous_Year),
               y= Total_Change_From_Previous_Year))+
    geom_bar(stat="identity", fill='steelblue') +
    coord_flip() +
    labs(title = "Unit Change from 2020 to 2021 by Company",
         x = "Company",
         y = "Unit Change From Previous Year") +
    theme(text = element_text(size = 6))
    

unit_change_from_previous_year
```
```{r}
# Plot highest sales by company in 2022
sales_by_company_2022 <- subset_QSR_2022 %>%
    ggplot(aes(x=reorder(Company, System_Sales_Millions), y=System_Sales_Millions))+
    geom_bar(stat="identity", fill='steelblue')+
    coord_flip()+
    labs(title='Sales in Millions by Company',
         x="Company",
         y="Sales (Millions)") +
    theme(text = element_text(size = 6))
sales_by_company_2022
```
In 2022, McDonalds had many more sales than other companies
```{r}
# Plot highest sales by company in 2021
sales_by_company_2021 <- subset_QSR_2021 %>%
    ggplot(aes(x=reorder(Company, System_Sales_Millions), y=System_Sales_Millions))+
    geom_bar(stat="identity", fill='steelblue')+
    coord_flip()+
    labs(title='Sales in Millions by Company',
         x="Company",
         y="Sales (Millions)") +
    theme(text = element_text(size = 6))
sales_by_company_2021
```
Again, McDonald's has the highest sales. For almost all companies, total sales in 2021 was lower than in 2022.
